---
title: "A3: SPL Library Checkouts"
output: html_document
---

### Introduction

This dataset informs us on book checkouts in the Seattle Public Library from 2013 up to 2022. It gives information on types of books, types of checkouts, authors, dates, publishers, etc. I chose to analyze the trends of physical book checkouts vs digital books over the years, the type of checkout someone did (magazines, music, movie, etc.) and the authors with the most books checked out. I did this because I am really interested in how the pandemic affected book checkouts and what kind of books people are interested in/checking out. I also chose to find the top authors because I am interested to know if I am familiar with any of them or their works, and I also wanted to see what people are reading the most.

### Summary Information
```{r, echo = FALSE, message = FALSE, code = readLines("summary.R")}
most_ebook_checkouts <- book_checkouts %>% 
  mutate("Date" = paste(sep = "-", CheckoutYear, CheckoutMonth, "01")) %>% 
  group_by(MaterialType, Date) %>% 
  filter(MaterialType == "EBOOK") %>% 
  summarise(most_ebook = max(Checkouts)) %>% 
  slice(which.max(most_ebook)) %>% 
  pull(Date)

least_ebook_checkouts <- book_checkouts %>% 
  mutate("Date" = paste(sep = "-", CheckoutYear, CheckoutMonth, "01")) %>% 
  group_by(MaterialType, Date) %>% 
  filter(MaterialType == "EBOOK") %>% 
  summarise(least_ebook = min(Checkouts)) %>% 
  slice(which.min(least_ebook)) %>% 
  pull(Date)

kate_dicamillo <- book_checkouts %>% 
  filter(Creator == "Kate DiCamillo") %>% 
  mutate("Date" = paste(sep = "-", CheckoutYear, CheckoutMonth, "01")) %>% 
  group_by(Date, Title) %>% 
  summarise(total_checkouts = sum(Checkouts)) %>% 
  arrange(desc(total_checkouts)) %>% 
  top_n(1) %>% 
  pull(Title) 

kate_most<- kate_dicamillo[1]

kate_dicamillo_checkout <- book_checkouts %>% 
  filter(Creator == "Kate DiCamillo") %>% 
  mutate("Date" = paste(sep = "-", CheckoutYear, CheckoutMonth, "01")) %>% 
  group_by(Date, Title) %>% 
  summarise(total_checkouts = sum(Checkouts)) %>% 
  arrange(desc(total_checkouts)) %>% 
  top_n(1) %>% 
  pull(total_checkouts) 

kate_top <- kate_dicamillo_checkout[1]

print_books_2013 <- book_checkouts %>% 
  group_by(UsageClass) %>% 
  filter(CheckoutYear == "2013") %>% 
  filter(UsageClass == "Physical") %>% 
  summarise(phys_2013 = sum(Checkouts)) %>% 
  pull(phys_2013)

dig_books_2013 <- book_checkouts %>% 
  group_by(UsageClass) %>% 
  filter(CheckoutYear == "2013") %>% 
  filter(UsageClass == "Digital") %>% 
  summarise(dig_2013 = sum(Checkouts)) %>% 
  pull(dig_2013)

perc_phys_2013 <- (print_books_2013 / (dig_books_2013 + print_books_2013)) * 100

print_books_2022 <- book_checkouts %>% 
  group_by(UsageClass) %>% 
  filter(CheckoutYear == "2022") %>% 
  filter(UsageClass == "Physical") %>% 
  summarise(phys_2022 = sum(Checkouts)) %>% 
  pull(phys_2022)

dig_books_2022 <- book_checkouts %>% 
  group_by(UsageClass) %>% 
  filter(CheckoutYear == "2022") %>% 
  filter(UsageClass == "Digital") %>% 
  summarise(dig_2022 = sum(Checkouts)) %>% 
  pull(dig_2022)

perc_phys_2022 <- (print_books_2022 / (dig_books_2022 + print_books_2022)) * 100

diff_physical <- print_books_2013 - print_books_2022

```

While analyzing the dataset, I found a few interesting values that reflect the trends I noticed. The first trend I found was confirmed by my first chart. The amount of Physical books being checked out has dramatically decreased between 2013 and 2022. In 2013, physical books made up `r perc_phys_2013` of all the books being checked out. However, in 2022 they only made up `r perc_phys_2022` of all books checked out. Considering the pandemic which didn't allow people to leave their house as much as usual, this number makes sense. Another trend that aligns with this fact is the amount of ebooks checked out in recent years. The date with the highest amount of ebook checkouts is `r most_ebook_checkouts`, unsurprisingly this date is in 2022. `r least_ebook_checkouts` is the date with the least amount of ebook checkouts, which is in 2013. This follows previous trends we have seen, since the popularity of digital books has increased over the past ten years. My favorite author, Kate DiCamillo, has a few books featured in the dataset, however her most popular book is `r kate_most`, with a high of `r kate_top` checkouts in a single month.

### The Dataset

- Who collected/published the data?  
- What are the parameters of the data (dates, number of checkouts, kinds of books, etc.)?
- How was the data collected or generated?  
- Why was the data collected?  
- What, if any, ethical questions do you need to consider when working with this data?  
- What are possible limitations or problems with this data?   (at least 200 words)


### Physcial VS Digital Book Checkouts

For my first trends over time chart, I decided to analyze checkouts of physical versus digital books over 2013-2023. I chose to analyze these variables because I knew the trend that would occur would be very interesting. I knew that the recent pandemic and technology becoming more accessible over the past decade will influence what kinds of books people are checking out. My suspicions turned out to be true because there is a very clear trend in both the digital and physical books being checked out. Physical books are being checked out less and less over time, while the opposite is happening to digital books.

```{r, echo = FALSE, message = FALSE, code = readLines("Chart1.R")}

physical_values <- book_checkouts %>% 
  group_by(CheckoutYear) %>% 
  filter(UsageClass == "Physical") %>% 
  summarise(avg_physical = mean(Checkouts))
digital_values <- book_checkouts %>% 
  group_by(CheckoutYear) %>% 
  filter(UsageClass == "Digital") %>% 
  summarise(avg_digital = mean(Checkouts)) 
book_type <- left_join(physical_values, digital_values, join_by("CheckoutYear"))


ggplot(book_type) +
  geom_line(mapping = aes(x = CheckoutYear, y = avg_physical, color = "Physical")) +
  geom_line(mapping = aes(x = CheckoutYear, y = avg_digital, color = "Digital")) +
  labs(title = "Physical VS Digital Book Checkout Over A Decade", x = "Year", y = "# of Books Checked Out Per Person", color = "Type of Book") + 
  scale_x_continuous(breaks = seq(2013, 2023, 2))
  
  
```

### Types of Checkouts over A Decade

I included this chart to show the trends of the type of checkouts over time. By type of checkout, it  means if people checked out books, movies, music, etc. Each checkout corresponds to what that person checked out, as labeled in the legend below. I chose to analyze this because I wanted to see what people like to check out of the library, and how peoples interests change over time. While most types of checkouts stayed in a steady trendline, Music seemed to decrease by quite a bit, which is explained by the introduction of iPhones and apps such as Spotify and Apple Music. A big trend I noticed was Zinio, which is how people can checkout Magazines and Newspapers online. It is the only trend actively increasing and it also started out very high. This could be explained because people mostly read on their digital devices, and this trend has only been increasing since 2013. 

```{r, echo = FALSE, message = FALSE, code = readLines("Chart2.R")}
freegal_checkouts <- book_checkouts %>% 
  group_by(CheckoutYear, CheckoutType) %>% 
  summarise(avg_freegal = mean(Checkouts)) %>% 
  filter(CheckoutType == "Freegal") %>% 
  select(CheckoutYear, avg_freegal)

hoopla_checkouts <- book_checkouts %>% 
  group_by(CheckoutYear, CheckoutType) %>% 
  summarise(avg_hoopla = mean(Checkouts)) %>% 
  filter(CheckoutType == "Hoopla") %>% 
  select(CheckoutYear, avg_hoopla)

horizon_checkouts <- book_checkouts %>% 
  group_by(CheckoutYear, CheckoutType) %>% 
  summarise(avg_horizon = mean(Checkouts)) %>% 
  filter(CheckoutType == "Horizon") %>% 
  select(CheckoutYear, avg_horizon)

overdrive_checkouts <- book_checkouts %>% 
  group_by(CheckoutYear, CheckoutType) %>% 
  summarise(avg_overdrive = mean(Checkouts)) %>% 
  filter(CheckoutType == "OverDrive") %>% 
  select(CheckoutYear, avg_overdrive)

zinio_checkouts <- book_checkouts %>% 
  group_by(CheckoutYear,CheckoutType) %>% 
  summarise(avg_zinio = mean(Checkouts)) %>% 
  filter(CheckoutType == "Zinio") %>% 
  select(CheckoutYear, avg_zinio)

all_checkouts <- freegal_checkouts %>% 
  left_join(hoopla_checkouts, by = "CheckoutYear") %>% 
  left_join(horizon_checkouts, by = "CheckoutYear") %>% 
  left_join(overdrive_checkouts, by = "CheckoutYear") %>% 
  left_join(zinio_checkouts, by = "CheckoutYear")

ggplot(all_checkouts) +
  geom_line(aes(x = CheckoutYear, y = avg_freegal, color = "Freegal (Music)")) +
  geom_line(aes(x = CheckoutYear, y = avg_hoopla, color = "Hoopla (Movies and TV)")) +
  geom_line(aes(x = CheckoutYear, y = avg_horizon, color = "Horizon (Physical Books)")) +
  geom_line(aes(x = CheckoutYear, y = avg_overdrive, color = "Overdrive (E-Books)")) +
  geom_line(aes(x = CheckoutYear, y = avg_zinio, color = "Zinio (Magazines and Newspapers)")) +
  labs(title = "Types of Checkout Over A Decade", x = "Year", y = "Average # of Things Checked Out Per Person", color = "Type of Checkout") + 
  scale_x_continuous(breaks = seq(2013, 2023, 2))


```

### Top Authors by Number of Checkouts

For the last chart, I wanted to look at what authors are the most popular and I measured that by looking at their overall checkouts. I specifically wanted to know about authors, not the publishing company, and I wanted to make sure the author was a single person and had their name updated in the data set that I used. I wanted to know if I was familiar with any of the most popular authors among Seattles Public Libraries. I found that the most popular author was Mo Willems, the author of many well known childrens books. The next most famous author is Jim Davis, who is the creator of Garfield. I found an interesting trend among the top 5 authors. They were all children's books authors, which may give a huge clue as to who is checking out the most books from Seattle Public Libraries.

```{r, echo = FALSE, message = FALSE, code = readLines("Chart3.R")}

most_common <- book_checkouts %>% 
  group_by(Publisher, Creator) %>% 
  filter(Creator != "") %>% 
  summarise(Checkouts = sum(Checkouts)) %>% 
  arrange(desc(Checkouts))

top_5 <- head(most_common, 5)

ggplot(top_5, aes(x = Creator, y = Checkouts, color = Creator, fill = Creator)) +
  geom_bar(stat = "identity")+
  labs(title = "Top Authors by Number of Checkouts")

```